\chapter{Background}
\label{chapterBackground}

\epigraph{{\it Although the authors work on formal grammars involving nontrivial TCS techniques, it is not clear that what there is a theoretical contribution or just common sense.}}{Reviewer \#2, [redacted]}
\epigraph{\it Much of the methodology of grammar testing is dictated by common sense.}{Miriam Butt et al.\\{\it A Grammar Writer's Cookbook}}

In this chapter, we present the main areas of this thesis: computational natural language grammars and software testing.
Section~\ref{sec:cg-intro} introduces the CG formalism and some of the different variants, along with the historical context and related work.
For a description of any specific CG implementation, we direct the reader to the original source: CG-1 \cite{karlsson1990cgp,karlsson1995constraint}, CG-2 \cite{tapanainen1996}, VISL CG-3 \cite{bick2015,vislcg3}.
Section~\ref{sec:gf-intro} introduces the GF formalism \cite{ranta2011gfbook} and the Resource Grammar Library \cite{ranta2009rgl}.

Section~\ref{sec:testing-intro} is a brief and high-level introduction to software testing and verification, aimed for a reader with no prior knowledge on the topic. For a more thorough introduction on software testing in general, we recommend \cite{ammann2016introduction}, and for SAT in particular, we recommend \cite{biere2009handbook}.




% \todo{quote from a LFG book---maybe find some use for it}
% \begin{quote}
% Testing the grammar is an indispensable part of ensuring robustness
% and increasing the performance of a grammar. Although individual constructions
% can be tested with a few sample sentences as they are developed,
% more systematic testing is required to maintain the quality of
% the grammar over time and to catch unexpected results of the addition
% of new rules and lexical items. Much of the methodology of grammar
% testing is dictated by common sense. However, issues do arise with respect
% to what kinds of testsuites to use, what other kinds of methods
% apart from testsuites one might use to test the grammar for continued
% consistency, and how to get a stable measure of the grammar's performance
% as it grows and changes.
% \end{quote}

\input{chapters/grammars-intro}

\input{chapters/testing-intro}


\section{Summary}

In this section, we have presented the theoretical background used in this thesis. 
We have introduced Constraint Grammar and Grammatical Framework as examples of computational grammars.
On the software testing side, we have presented key concepts such as unit testing and 
Boolean satisfiability. %: methods for solving whether a given propositional formula can evaluate to true.
In the following chapters, we will connect the two branches. Firstly, we encode CG grammars as SAT-problems, which allows us to apply symbolic evaluation and find potential conflicts between rules. 
Secondly, we use methods for creating minimal and representative test data, in order to find a set of trees that test a given GF grammar in an optimal way.



